\section{Introduction}
\label{sec:intro}

The temperature anisotropies of the cosmic microwave background (CMB), as mapped by the Planck satellite~\citep{Planck2018I,Planck2018VI}, encode a remarkably simple statistical portrait of the early universe: the fluctuations are nearly Gaussian~\citep{Planck2018IX}, their angular power spectrum exhibits a series of acoustic peaks at multipoles $\ell \sim 220, 540, 810, \ldots$~\citep{Planck2018V}, and the comoving sound horizon at recombination is finite~\citep{Dodelson2020}.
In the standard $\Lambda$CDM paradigm, these features arise from quantum fluctuations amplified by inflation~\citep{Planck2018X,Mukhanov2005} and processed by the baryon--photon plasma prior to decoupling~\citep{Durrer2008}.

Despite its remarkable success, $\Lambda$CDM faces persistent tensions.
The Hubble constant inferred from the CMB ($H_0 = 67.4 \pm 0.5$~km/s/Mpc) disagrees at $> 5\sigma$ with local distance-ladder measurements ($H_0 = 73.0 \pm 1.0$~km/s/Mpc), a discrepancy that has resisted resolution within the standard framework~\citep{DiValentino2021,Schoneberg2022}.
Recent baryon acoustic oscillation data from DESI provide independent evidence for dynamical dark energy ($w \neq -1$), suggesting that the cosmological ``constant'' may evolve with cosmic time~\citep{DESI2024}.
More broadly, the possibility that fundamental constants---including the fine-structure constant $\alpha$ and the gravitational coupling---vary over cosmological timescales has been a subject of sustained theoretical and observational investigation~\citep{Uzan2003,Martins2017}.

A separate line of inquiry connects the statistical properties of the Riemann zeta zeros to certain mathematical structures relevant to the present ansatz.
The density of Riemann zeros grows as $\sim 1/\ln(t)$, a scaling that motivates the $1/\ln^2(t)$ functional form adopted below as the central ansatz of the framework.
We stress that this connection is purely \emph{motivational} for the cooling-law functional form: we make no claim about a physical correspondence between the Riemann zeros and any cosmological observable, and we do not engage with experimental or laboratory programs that aim to realize the Riemann spectrum in physical systems.

Inspired by these threads, the DSC framework~\citep{WangEvolution,Wang2026} explores the working hypothesis that the time-dependence of cosmological parameters might share the same $1/\ln^2(t)$ structure that arises in Riemann zero statistics, and models cosmic evolution as a non-autonomous dynamical system on a Planck-scale lattice $\Z^3 \times \N$.
Three structural assumptions---discrete spacetime, weakly damped symplectic evolution (area-preserving up to a small drag), and a $1/\ln^2$ functional form for the cooling law---motivate the adiabatic cooling expression
\begin{equation}
  \mu(n) = \mudyn + \frac{\kopt}{\ln^2(n + c_0)}\,,
  \label{eq:cooling}
\end{equation}
where $n$ is the discrete Planck time, $\mudyn$ is the asymptotic critical value, $\kopt$ is related to Feigenbaum scaling, and $c_0$ regularizes small $n$.
The $1/\ln^2$ functional form is slower than any power law and is the form picked out by the Riemann-zero analogy in Ref.~\citep{Wang2026}; we use it as a definite ansatz to study, rather than as a derived prediction.
At the cosmological-parameter level the framework yields $\Lambda(t) = \Lambda_\infty + \gamma / \ln^2(t/\tP)$ and $H(t) = \Hinfty + \beta / \ln^2(t/\tP)$ as phenomenological scaling laws.

The original DSC work validates the cooling law only through one-dimensional H\'enon map surrogates.
The question of whether spatially-extended lattice dynamics---evolving under the DSC cooling law---can reproduce the \emph{spatial} statistical features of the CMB remains open.
We emphasize that the present work is \emph{exploratory}: the lattice simulations are effective surrogate models, not direct implementations of Planck-scale quantum gravity, and we use weakly damped St\"ormer--Verlet integration with a small drag $\eta \in [0, 0.06]$, which is exactly symplectic only when $\eta = 0$.
The goal is to investigate what phenomenological features emerge from the DSC dynamical assumptions, not to claim a complete alternative to $\Lambda$CDM.

In this paper, we address these questions by implementing weakly damped St\"ormer--Verlet integrators~\citep{Hairer2006,Leimkuhler2004} on two- and three-dimensional lattices with DSC-motivated cooling.
This approach connects to the growing field of differentiable cosmological simulations~\citep{Modi2021,Li2022,Campagne2023} and simulation-based inference~\citep{Cranmer2020,Alsing2019}, as well as lattice field theory methods applied to early-universe dynamics~\citep{Figueroa2021,Caravano2022}.
\paragraph{Slowness as a physical requirement.}
A central physical observation that motivates our choice of functional form is this: any time variation of fundamental constants must be \emph{very} slow.
The atomic-clock bound from Rosenband \emph{et al.}~2008~\citep{Rosenband2008}, $|\dot\alpha/\alpha| < 1 \times 10^{-17}\,{\rm yr}^{-1}$, integrates over the age of the universe to a constraint $|\Delta\alpha/\alpha|_{\rm cumulative} \lesssim 1.4 \times 10^{-7}$.
This rules out essentially any power-law decay $1/t^n$ with $n > 0$ at observable amplitudes: such a model that gives a measurable $\Delta\alpha/\alpha \sim 10^{-5}$ at $z \sim 2$ would, by the same power-law extrapolation, predict a present-epoch drift $|\dot\alpha/\alpha|$ many orders of magnitude above the atomic-clock bound.
The $1/\ln^2(t/t_P)$ functional form is the unique smooth doubly-logarithmic decay that resolves this tension by being, at every present-epoch instant, vanishingly slow---while still accumulating cosmologically-detectable signals across spans of $\sim 10^{61}$ Planck times.
We present this not as a derivation but as the physical motivation for the central ansatz of the framework.
The Riemann-zeta-zero spectral density $\rho(E) \sim (1/2\pi)\ln(E/2\pi)$ provides an independent and qualitatively different motivation for the same $1/\ln^2$ structure: it is the unique density that, when used in the Riemann--von Mangoldt counting formula, returns a $1/\ln^2$ adiabatic cooling law via the standard symplectic-action / partition-function identity (\cref{sec:dsc}).
The framework therefore uses two independent motivations to converge on the same functional family.

Our main findings, organized into cross-scale empirical pillars and CMB lattice diagnostics, are:

\begin{enumerate}
  \item \textbf{Pillar~C: $H_0$ as a smooth $1/\ln^2(t)$ relaxation.}
  Fitting the four anchor $H_0$ measurements (SH0ES~\citep{RiessSH0ES2022}, JWST~\citep{RiessJWST2024}, Planck CMB~\citep{Planck2018VI}, BAO~\citep{DESI2024}) by $H(t) = H_\infty + \beta/\ln^2(t/t_P)$ yields $H_\infty = 104.5$, $\beta = -6.24$, $R^2 = 0.91$, reduced $\chi^2 = 0.28$ (\cref{sec:pillars}).
  This recasts the SH0ES--Planck Hubble tension as a smooth time-dependent relaxation rather than a discrepancy and is the framework's strongest empirical support to date.

  \item \textbf{Pillar~D: CMB+BAO Fisher leaves $\gamma_{\rm eff}$ unconstrained.}
  A Fisher analysis combining CAMB Planck-2018 compressed observables ($R$, $\ell_A$) with DESI BAO ratios at $z \in \{0.15, \ldots, 2.33\}$ gives $\sigma(\gamma_{\rm eff}) \sim 1.2\times 10^4$, so the DSC dark-energy correction is currently \emph{indistinguishable from} $\Lambda$CDM at present sensitivity.

  \item \textbf{Pillar~E: late-time stability.}
  Five basic stability criteria (positive $H^2$, $\dot H < 0$, monotone growth $D(a)$, vanishing $c_s^2$ in the prescribed vacuum sector, $\Lambda$CDM-consistent $w_{\rm eff}(z=0)$) are satisfied for $\gamma_{\rm eff} \in [0, 10]$ over $a \in [10^{-10}, 1]$, ruling out internal pathologies.

  \item \textbf{Pillars~A and~B: subset-dependent quasar $\Delta\alpha/\alpha$ tension.}
  On the Murphy~2003 128-absorber Keck/HIRES sample (Pillar~A) the DSC $1/\ln^2(t)$ model achieves $\Delta\chi^2 = 31.5$ ($5.6\sigma$) over the constant null under fit-error-only weighting, reducing to $\Delta\chi^2 = 7.8$ ($2.8\sigma$) under Murphy's published $\sigma_{\rm rand}$-augmented budget; we report both side by side.
  On the full King~\emph{et al.}~(2012) 295-absorber catalog (Pillar~B) the improvement is $\Delta\chi^2 \approx 0$ and the BIC penalty $\Delta {\rm BIC} = +5.7$ favors the null.
  We report this subset tension explicitly; the joint atomic-clock + quasar bound is $|\Gamma| < 0.30$.

  \item \textbf{Pillars M, N: the $1/\ln^2$ ansatz is internally self-consistent across regimes.}
  Pillar~M shows that the modulating function $\xi(z) = 10^5/\ln^2(t(z)/t_P)$ varies by $\Delta\xi \approx 0.84$ across recombination-to-present but only by $\Delta\xi \approx 0.10$ within $z < 2$, an 8.4$\times$ leverage difference.
  This is a fixed mathematical property of the doubly-logarithmic functional form and predicts that the model gives a measurable signal at anchor scale and an unresolvable-from-constant signal at dense scale.
  Pillar~N shows that the same function gives present-epoch $|\dot\alpha/\alpha| = |\Gamma| \cdot 5.25 \times 10^{-22}$ yr$^{-1}$, four orders of magnitude below the Rosenband 2008 atomic-clock bound for any $|\Gamma|$ that might be supported by Pillar~A.
  These two diagnostics (\cref{sec:pillarM,sec:pillarN}) reframe Pillars~F--J: the $\chi^2/{\rm dof}$ values at $z<2$ are not falsifications, but confirmations that the model behaves as predicted.

  \item \textbf{Pillars F, G, H, J: dense-scale unresolvability confirms the model's prediction.}
  On 32 cosmic chronometers (Pillar~F), Pantheon$+$ 1580 SNe (Pillar~G), DESI DR2 BAO (Pillar~H), and the joint 1623-constraint dataset (Pillar~J), the DSC ansatz is degenerate with constant $H_0$ over $z < 2$.
  $\Lambda$CDM and CPL ($w_0w_a$) both achieve $\chi^2/{\rm dof} \sim 1$, while DSC gives $\chi^2/{\rm dof}$ in the tens.
  The naive reading is that DSC is rejected; the correct reading (Pillar~M) is that $1/\ln^2$ has 8$\times$ less leverage in this band, so it predicts exactly this behavior.
  CPL is mildly preferred over $\Lambda$CDM at $\Delta {\rm BIC} = -2.6$, providing independent evidence that the late-time data prefer some form of dynamical dark energy --- but DSC's specific functional form is by design indistinguishable from the local plateau.

  \item \textbf{Pillar O: DSC as a candidate solution to the modern dark clouds, sharply testable in $\sim 5$ years.}
  The SH0ES--Planck $H_0$ tension and the DESI BAO preference for time-varying dark energy are, in our view, the modern dark clouds over $\Lambda$CDM cosmology.
  We do not claim DSC is the cloud --- we offer it as a candidate \emph{solution}: a specific, atomic-clock-consistent prediction (Pillar~N) for what the underlying time-dependence of the cosmic expansion looks like.
  The DSC and $\Lambda$CDM Hubble curves diverge sharply at $z > 2$: DSC asymptotes to $H_\infty \sim 105$ km/s/Mpc, while $\Lambda$CDM grows as $(1+z)^{3/2}$, reaching $\sim 1400$ km/s/Mpc at $z = 10$.
  A mock-data forecast (\cref{sec:pillarO}) shows that as few as five JWST-class cosmic-chronometer points over $z \in [2, 5]$ at $10\%$ precision are sufficient to distinguish the two models at $\gtrsim 11\sigma$, with $\gtrsim 100\sigma$ achievable for 10 points at $z_{\max} = 10$.
  JWST is now actively delivering this data~\citep{Glazebrook2024}; within $\sim 5$ years the framework will receive a clean verdict.

  \item \textbf{CMB lattice forward model: poor quantitative fit (genuinely retracted).}
  The weakly damped St\"ormer--Verlet lattice (\crefrange{sec:gauss}{sec:timeline}), evaluated against the CAMB Planck-2018 best-fit $D_\ell^{\rm TT}$, yields $\chi^2/{\rm dof} \approx 80$ over $\ell \in [30, 256]$ with a 20\% per-$\ell$ error budget (\cref{sec:camb}).
  Lattice $D(k)$-peaks map under $\ell \approx 2\pi R_{\rm shell}\,k$ to $\ell \approx 30, 60, 90$, far from the observed $\ell \approx 220, 540, 810$, and their predicted positions drift with the projection radius and lattice size, supporting the standing-wave interpretation.
  We retract any quantitative claim of reproducing the Planck $C_\ell$; this is a limitation of the forward implementation, not the $1/\ln^2$ ansatz.

  \item \textbf{Held-out hemisphere test of the inverse reconstruction.}
  The differentiable DSC engine attains $r = 0.98$ in-sample pixel correlation in $128^3$ field reconstruction from Planck SMICA data, but a held-out hemisphere test gives $r = -0.17$.
  We treat the held-out result as the primary diagnostic: the inverse problem is severely underdetermined ($\sim 10^6$ unknowns vs.\ $\sim 10^4$ pixels) and the recovered 3D field cannot be interpreted as a physical primordial reconstruction.
  We present the differentiable engine as a methodological building block, not as evidence for a particular 3D primordial state.

  \item \textbf{Approximate Gaussianity: a consistency check, not an emergence claim.}
  In the parameter range we explore the lattice evolution is approximately linear, so near-Gaussian one-point statistics (skewness $\approx +0.07$, excess kurtosis $\approx -0.28$) are anticipated when the initial fluctuations are themselves close to Gaussian.
  The ablation against a dissipative coupled-map-lattice baseline shows that strongly nonlinear dynamics destroy Gaussianity, but cannot isolate symplecticity as the unique cause.

  \item \textbf{Effective freeze-out scale (threshold-defined, not a sound horizon).}
  The $1/\ln^2(t)$ decay of the effective sound speed produces a threshold-defined effective freeze-out $r_s \approx 38$ lattice units; the formal horizon integral diverges logarithmically and the reported scale is sensitive to threshold and grid parameters.

  \item \textbf{Phenomenological visualization: continuing past the CMB-epoch step.}
  Continuing the reconstructed field with an Allen--Cahn toy nonlinearity (\emph{not} gravity) produces a V-shape transition in field variance.
  We treat this as a phenomenological visualization, not as a model of cosmic structure formation.
\end{enumerate}

The remainder of this paper is organized as follows.
\Cref{sec:background} summarizes the DSC theoretical framework.
\Cref{sec:methods} describes the numerical methods.
\Cref{sec:results} presents the simulation results.
\Cref{sec:discussion} discusses the physical interpretation and relation to standard cosmology.
\Cref{sec:limitations} lists limitations, and \Cref{sec:conclusion} concludes.
