\section{Discussion}
\label{sec:discussion}

\paragraph{What the Gaussianity finding does and does not show.}
The near-Gaussian one-point statistics of the lattice are consistent with two simple expectations rather than a non-trivial emergence.
First, the initial conditions are random Gaussian fields, and the St\"ormer--Verlet update is linear in $\phi$ in the parameter range we explore; a linear evolution of a Gaussian field is also Gaussian.
Second, each pixel of the projected sky is a sum over many lattice modes, and the central limit theorem favors Gaussianity regardless of the individual mode statistics.
The contrast with the dissipative CML baseline (kurtosis $= -1.60$) shows that strongly nonlinear dynamics destroy this Gaussianity, but the comparison cannot isolate symplecticity as the active ingredient: the two baselines differ in nonlinearity, conservation properties, and integration order simultaneously.

\paragraph{Lattice peaks vs.\ baryon--photon acoustic oscillations.}
The peaks in the lattice $D(k)$ arise from standing-wave physics on a finite-volume lattice with a time-dependent effective propagation speed.
On a sphere of radius $R$ inside the lattice they project, in the flat-sky limit, to multipoles $\ell \approx 2\pi R_{\rm shell}\,k$.
With the geometry we use this gives $\ell \approx 30, 60, 90$ for the first three lattice peaks, far from the Planck multipoles $\ell \approx 220, 540, 810$.
The peak positions in $\ell$-space drift with the projection radius and the lattice size (\cref{sec:camb}), which is the behavior expected from standing waves rather than from physical acoustic oscillations.
We therefore retract the earlier framing of these peaks as analogs of baryon--photon acoustic oscillations.

\paragraph{Status of the connection to the DSC framework.}
The simulations here implement the cooling law of Ref.~\citep{Wang2026} as a definite ansatz to study, not as a derived prediction of Planck-scale physics.
The motivation from Riemann zero spectral statistics is treated here as a working motivation for the $1/\ln^2$ functional form of the cooling law, not as an established physical correspondence.
We do not engage with the broader question of whether a physical Hamiltonian exists whose spectrum reproduces the Riemann zeros, nor with laboratory programs aimed at realizing such a spectrum; those questions are outside the scope of the present manuscript.
Within the lattice surrogate we observe that the $1/\ln^2(t)$ cooling controls the location of the threshold-defined effective freeze-out scale (\cref{sec:horizon}), but this scale is sensitive to threshold and grid parameters and we do not interpret it as an analog of the comoving sound horizon.

\paragraph{Comparison with related approaches.}
Coupled map lattices have been studied as models for spatiotemporal chaos and pattern formation~\citep{Kaneko1992}, but not in a cosmological context with adiabatic cooling.
Lattice cosmology approaches~\citep{Regge1961} discretize Einstein's equations directly, while causal set theory~\citep{Bombelli1987} replaces the manifold with a partial order.
The DSC lattice differs from both: it does not attempt to recover general relativity from the lattice, and it should not be read as an alternative formulation of cosmology.
The differentiable simulation-based inference component connects to the growing field of likelihood-free inference in cosmology~\citep{Alsing2019,Cranmer2020}, where forward simulations replace analytical likelihoods, and to differentiable cosmological simulators~\citep{Modi2021,Li2022,Campagne2023}.

\paragraph{Comparison to existing time-varying $H_0$ analyses, and the anchor-vs-dense leverage asymmetry.}
A small but persistent line of work has searched directly for $H_0(z)$ trends in cosmological data.
Dainotti~\emph{et al.}~2021~\citep{Dainotti2021} binned the Pantheon Type Ia supernova compilation in redshift and reported a $\sim 2\sigma$ descending trend $H_0(z) = \tilde H_0/(1+z)^\alpha$ with $\alpha \approx 0.01$.
Krishnan~\emph{et al.}~2020~\citep{Krishnan2020} found a similar mild descending trend combining Pantheon, BAO, and cosmic-chronometer data ($\sim 2\sigma$).
Kazantzidis \& Perivolaropoulos~2020~\citep{Kazantzidis2020} report $\sim 2\sigma$ on Cepheid+TRGB distance ladders.
Colg{\'a}in~\emph{et al.}~2022~\citep{Colgain2022} extended the analysis with quasar+BAO data and again obtained $\lesssim 2\sigma$.
The Hubble-tension model census of Schöneberg~\emph{et al.}~2022~\citep{Schoneberg2022} ranks dozens of solutions (early dark energy, modified recombination, etc.) and notes that none currently survive joint multi-probe scrutiny.

Three observations follow from comparing this literature with the present work.

(i)~\emph{Functional form.} Existing analyses adopt power-law $(1+z)^\alpha$ ansätze; we do not know of any prior published cosmological work fitting a $1/\ln^2(t/t_P)$ Hubble relaxation to $H_0$ anchors.
The functional form here is independent of the literature ansätze, motivated by the Riemann-zeta-zero spectral density (\cref{sec:dsc}) and by the slowness requirement of \cref{sec:pillarN}.

(ii)~\emph{Leverage choice.} The literature analyses operate exclusively within the $z < 2.3$ band of Pantheon (or the comparable bands of BAO and CC), where Pillar~M shows the $1/\ln^2$ ansatz has $\Delta\xi \approx 0.10$ of dynamic range.
The mild $\sim 2\sigma$ trends those analyses report are therefore consistent with what the $1/\ln^2$ model predicts: a barely-resolvable shape over $z < 2$.
By using the four-anchor recombination-to-present span (Pillar~C), we operate in the $\Delta\xi \approx 0.84$ regime where the same ansatz has 8.4$\times$ the leverage---hence $R^2 = 0.91, \chi^2/{\rm dof} = 0.28$ rather than $\sim 2\sigma$.
The two analyses are not in tension; they are different leverage slices of the same underlying functional family.

(iii)~\emph{The anchor-vs-dense asymmetry is a generic property of the field.}
The same asymmetry appears in essentially every modern Hubble-tension proposal.
Early Dark Energy~\citep{Schoneberg2022} resolves the SH0ES--Planck anchor difference by modifying recombination physics, but is in tension with low-redshift growth-rate measurements ($\sigma_8$ tension; the so-called yin-yang problem).
CPL ($w_0w_a$) achieves a marginal $\Delta {\rm BIC} = -2.6$ on dense BAO+SN+CC data (Pillar~J) but does not address the anchor-level $H_0$ tension by construction.
Modified-recombination scenarios match Planck but introduce tensions elsewhere.
The DSC $1/\ln^2$ ansatz, like all of these proposals, exhibits a leverage-dependent fit quality.
What is, to our knowledge, novel here is making the asymmetry mathematically explicit (Pillar~M, 8.4$\times$ ratio fixed by the doubly-logarithmic form), bounding it by laboratory atomic-clock measurements (Pillar~N), and providing a quantitative falsification roadmap (Pillar~O, $\gtrsim 11\sigma$ discrimination from $\sim 5$ JWST cosmic-chronometer points at $z \in [2,5]$).
Whether this self-consistency framing extends to other modern proposals (e.g.\ EDE) is, in our view, a worthwhile direction for the field.

\paragraph{Forward identifiability vs.\ field-level underdetermination.}
The forward twin experiment (\cref{sec:twin}) optimizes 3 global physics parameters against $\sim 75$ power-spectrum bins, an overdetermined system that recovers the parameters up to a $c^2$--$\kopt$ degeneracy direction.
The voxel-level reconstruction (\cref{sec:inverse}), in contrast, attempts to recover $\sim 10^6$ free parameters from $\sim 10^4$ sky pixels.
The high in-sample correlation $r = 0.98$ is not surprising at this ratio of unknowns to data, and the held-out hemisphere result $r = -0.17$ is the appropriate diagnostic: the inverse problem is severely underdetermined and the recovered 3D field cannot be interpreted as a physical primordial reconstruction.
We therefore present the differentiable engine---gradients flowing cleanly through 25 St\"ormer--Verlet steps and the sphere-projection layer---as a methodological building block for future multi-probe inference workflows that pair lattice forward models with volumetric priors, rather than as a tool for inferring the 3D primordial state from CMB data alone.

\paragraph{The continuation past the CMB-epoch step is a phenomenological visualization.}
The V-shape transition in field variance (\cref{sec:timeline}) is produced by a $1/\ln^2(t)$ cooling phase followed by an Allen--Cahn coarsening phase.
The Allen--Cahn nonlinearity is a generic toy potential, not a gravitational interaction, and the resulting domain morphology is generic to coarsening dynamics.
We retract the earlier suggestion of an analogy with cosmic inflation: the smoothing in our run is driven by an effective sound-speed decay rather than by a slow-roll scalar field, and the analogy is not quantitative.
The cosmic-timeline figure should be read as an illustration of the lattice's behavior under a chosen toy nonlinearity, not as a model of cosmic structure formation.
